---
title: "Exercise 5, Markup Languages"
subtitle: 'Replication study' 
author: "Naomi Smorenburg"
date: "12/1/2019"
output: html_document
---

  ```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 0. Start by setting a seed value
```{r}
set.seed(456)
```


# 1. Samples from a normal distribution 
Sample 100 samples from a standard normal distribution.
Replicate this study 100 times. 

```{r}
n <- 1000
mean <- 0
sd <- 1
samples <- replicate(100, rnorm(n, mean = mean, sd = sd))
```



# 2. Statistics 
For each of these samples, calculate the following statistics for the mean

Absolute bias:
  
  $$|\bar{x} - \mu| $$
  
  
Standard error: 
  
  $$\sigma_\bar{x}  = \frac{\sigma}{\sqrt{n}}$$
  
  
Where $$\sigma = \sqrt{\frac{1}{n} \Sigma(x_i - \mu)^2}$$
  
  
  Lower and upper bound of the 95% confidence interval: 
  
  $$ \bar{x} \pm 1.96 \frac{\sigma}{\sqrt{n}} $$
  
  ```{r}

# Function to calculate these statistics
info <- function(x){
  m <- mean(x)
  sem <- sd(x)/sqrt(n)
  output <- c(
    m,
    abs(m - mean),
    sem,
    m - 1.96*sem,
    m + 1.96*sem)
  
  return(output)
}

results2 <- t(apply(samples, 2, info))
colnames(results2) <- c("Mean", "bias", "sem", "lowerbound", "upperbound")
results2 <- data.frame(results2)

# add variable on type1 errors
library(magrittr)
results2 <- results2 %>%
  dplyr::mutate(typeone = mean > lowerbound & mean < upperbound)
```

# 3. Plot
Create a plot that demonstrates the following: 

A replication of the procedure that generates a 95% confidence interval that is centered around the sample mean would cover the population value at  least 95 out of 100 times" [(Neyman, 1934)](https://www.google.nl/url?sa=t&rct=j&q=&esrc=s&source=web&cd=1&ved=0ahUKEwjk4LaD6LrXAhUGC8AKHdATBhMQFggoMAA&url=http%3A%2F%2Fwww.stat.cmu.edu%2F~brian%2F905-2008%2Fpapers%2Fneyman-1934-jrss.pdf&usg=AOvVaw3KRw9DSklPihrYIM0_zRrC)

```{r}
library(ggplot2)
ggplot(data = results2, aes(x = 1:100, y = Mean, color = typeone)) + 
  geom_pointrange(aes(ymin = lowerbound, ymax = upperbound)) +
  xlab("Sample mean + 95% Confidence Interval") +
  ylab("Samples 1:100") +
  theme_classic()
```


# 4. Table
Present a table containing all simulated samples for which the resulting confidence interval does 
not contain the population value.

```{r}
# Load the following packages 
library(knitr)
library(kableExtra)
kable(results2[!results2$typeone, -6], caption = "Samples containing a Type 1 error") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

