---
title: "First Simulations"
author: "Naomi Smorenburg"
date: "November 2019"
output:
  html_document:
    df_print: paged
subtitle: Maximizing the expected overall life expectancy
---

Utility function maximizing the overall life expectancy for the entire population for a combination of values for t1 and t2. 

```{r}
# OLE = overall life expectancy 

OLE <- function(N, hE, hC, t1, t2, t_star) { 
          
      ### Function input ---
  
         # N = patient horizon 
         # hE = hazard rate experimental treatment 
         # hC = hazard rate control treatment 
         # t1 = starting point of trial (in years)
         # t2 = endpoint of trial (in years)
         # t_star = time until N new onset  
  
      ### Computations ---
  
  # 1. Calculate the sample size of the different groups
  n1 <- 0.5*N*(t1/t_star)           # Patients in the trial receiving the standard treatment  
  n2 <- 0.5*N*(t1/t_star)           # Patients in the trial receiving the experimental treatment
  n3 <- N*((t2-t1)/t_star)          # Patients not in the trial, with disease onset between t1 and t2
  n4 <- N*((t_star-t2)/t_star)      # Patients not in the trial, with disease onset after t2. 
  
  # 2. Calculating the average life expectancy when E is not superior 
  C.LE1 <- 1/hC                                                                   # group only receives C 
  C.LE2 <- (1/hE^2) * (hE + (((exp(t1*hE)-1)*exp(-t2*hE)*(hE-hC))/(t1*hC)))       # group receives E first, after t' C
  C.LE3 <- 1/hC                                                                   # group only receives C
  C.LE4 <- 1/hC                                                                   # group only receives C
    
  # 3. Calculating the average life expectancy when E is superior
  E.LE1 <- (1/hC^2) * (hC + (((exp(t1*hC)-1)*exp(-t2*hC)*(hC-hE))/(t1*hE)))       # group receives C first, after t' E                                                       
  E.LE2 <- 1/hE                                                                   # group only receives E
  E.LE3 <- (1/hC^2) * (hC + (((exp(t1*hC)-1)*exp(-t2*hC)*(hC-hE))/(t1*hE)))       # group receives C first, after t' E                                                      
  E.LE4 <- 1/hE                                                                   # group only receives E
             
  # 4. Calculating the overall life expectancy for the entire population N for situation E is superior and E is not          superior
  
        #### By multiplying group size with LE for individuals 
        C.OLE <- (n1 * C.LE1) + (n2 * C.LE2) + (n3 * C.LE3) + (n4 * C.LE4) 
        E.OLE <- (n1 * E.LE1) + (n2 * E.LE2) + (n3 * E.LE3) + (n4 * E.LE4) 
  
  # 5. Calculating the power 
  
        ### First calculate the expected number of events for each group 
       D.E <- n2 * (1 + ((exp(-1*(t2-t1)*hE) - exp(-1*t2*hE))/((t2-t1)*hE - t2*hE)))
       D.C <- n1 * (1 + ((exp(-1*(t2-t1)*hC) - exp(-1*t2*hC))/((t2-t1)*hC - t2*hC)))
 
       ### Calculate the total expected number of events by adding  both groups 
       D <- D.E + D.C
  
       #### PropE = proportion in E group, in the case of 1:1 allocation this is 0.5
       PropE <- 0.5
  
       ### Calculate the mean under the alternative hypothesis 
       ### Log-rank test statistic T has distribution (Schoenfeld, 1981 Biometrika)
       muHa <- log(hE/hC) * sqrt(PropE * (1-PropE) * D)
  
      ### Determine the critical value for one sided alpha of 0.05. 
      Cr <- qnorm(0.025, 0, 1) 
  
      #### Determine the power by taking the area left of the critical value 
      pwr <- pnorm(Cr, muHa, 1)
  
      # 6. Weigh the two possible outcomes by the power 
      U <- (pwr*E.OLE) + ((1-pwr)*C.OLE) 
  
      
      ### Output --- 
  
  # 7. Take the results and put in a list 
     result <- list(
    'Overall life expectancy E not superior' = C.OLE,
    'Overall life expectancy E superior' = E.OLE,
    'Utility' = U
    )
  
  # 6. Return a list with the results 
  return(result)

  # End of function 
}
```


First simulations
  Concerning the choice of a combination of t1 and t2
  Maximum duration of trial is 10 years (t_star = 10)
  t2 > t1 & t1 > 0 
  
```{r}
# 1. Create sequence with possible values for t1 and t2
# t_star stays contant = 10. 
t_star <-10

t1.seq <- seq(0, t_star, 0.1)
t2.seq <- seq(0, t_star, 0.1)
```


# Different scenario's 

# Scenario 1 
Experimental treatment is better then the control treatment (thus, lower hazard).
Only a small difference. N is relatively large.  
```{r}
# Data input
hE <-0.2
hC <- 0.25
N <- 3000

# Create matrix with possible combinations of t1 and t2
m <- expand.grid(hE=hE, hC=hC, t_star=t_star, N=N, t1=t1.seq, t2=t2.seq)    # t2 = column name, t2seq = content
m <- m[m$t2 >= m$t1 & m$t1>0,]                                              # add conditions: end of follow up (t2) needs                                                              to be after start of trial (t1), t1 must be larger than 0. 

# Add extra columns to matrix for C.OLE, E.OLE and Utility (U)
m$uc <- NA # 'Utility' C
m$ue <- NA # 'Utility' E
m$U <- NA # Overall utility weighted by the power 

# For loop
for(i in 1:nrow(m)) {
  m$uc[i] <- OLE(N = m$N[i], hE = m$hE[i], hC = m$hC[i], t1 = m$t1[i], t2 = m$t2[i], t_star = m$t_star[i])$'Overall life expectancy E not superior'
m$ue[i] <- OLE(N = m$N[i], hE = m$hE[i], hC = m$hC[i], t1 = m$t1[i], t2 = m$t2[i], t_star = m$t_star[i])$'Overall life expectancy E superior'
  m$U[i] <- OLE(N = m$N[i], hE = m$hE[i], hC = m$hC[i], t1 = m$t1[i], t2 = m$t2[i], t_star = m$t_star[i])$'Utility'
}

m

# Outcome 
m[m$U == max(m$U),]    
```


Plots
```{r}
library(rgl)

    # grey = E is not superior
    # green = E is superior
    # red = utility 

# All 3 combined
plot3d(x = m$t1, y = m$t2, z = m$uc, col = 'grey') + plot3d(x = m$t1, y = m$t2, z = m$ue, add = T, col = 'green') + plot3d(x = m$t1, y = m$t2, z = m$U, add = T, col = 'red')
 
# Only utility
plot3d(x = m$t1, y = m$t2, z = m$U, col = 'red')
```



# Scenario 2
Experimental treatment is better then the control treatment (thus, lower hazard).
A larger difference. N is relatively large. 
```{r}
# Data input
hE <-0.2
hC <- 0.4
t_star <-10
N <- 3000

# Create matrix with possible combinations of t1 and t2
m <- expand.grid(hE=hE, hC=hC, t_star=t_star, N=N, t1=t1.seq, t2=t2.seq)    # t2 = column name, t2seq = content
m <- m[m$t2 >= m$t1 & m$t1>0,]                                              # add conditions: end of follow up (t2) needs                                                              to be after start of trial (t1), t1 must be larger than 0. 

# Add extra columns to matrix for C.OLE, E.OLE and Utility (U)
m$uc <- NA # 'Utility' C
m$ue <- NA # 'Utility' E
m$U <- NA # Overall utility weighted by the power 

# For loop
for(i in 1:nrow(m)) {
  m$uc[i] <- OLE(N = m$N[i], hE = m$hE[i], hC = m$hC[i], t1 = m$t1[i], t2 = m$t2[i], t_star = m$t_star[i])$'Overall life expectancy E not superior'
m$ue[i] <- OLE(N = m$N[i], hE = m$hE[i], hC = m$hC[i], t1 = m$t1[i], t2 = m$t2[i], t_star = m$t_star[i])$'Overall life expectancy E superior'
  m$U[i] <- OLE(N = m$N[i], hE = m$hE[i], hC = m$hC[i], t1 = m$t1[i], t2 = m$t2[i], t_star = m$t_star[i])$'Utility'
}

# Outcome 
m[m$U == max(m$U),]  
```

# Plots 
```{r}
    # grey = E is not superior
    # green = E is superior
    # red = utility 

# All 3 combined
plot3d(x = m$t1, y = m$t2, z = m$uc, col = 'grey') + plot3d(x = m$t1, y = m$t2, z = m$ue, add = T, col = 'green') + plot3d(x = m$t1, y = m$t2, z = m$U, add = T, col = 'red')
 
# Only utility
plot3d(x = m$t1, y = m$t2, z = m$U, col = 'red')
```



# Scenario 3 
Experimental treatment is better then the control treatment (thus, lower hazard).
A larger difference. N is relatively small (300). 
```{r}
# Data input
hE <-0.2
hC <- 0.4
t_star <-10
N <- 300

# Create matrix with possible combinations of t1 and t2
m <- expand.grid(hE=hE, hC=hC, t_star=t_star, N=N, t1=t1.seq, t2=t2.seq)    # t2 = column name, t2seq = content
m <- m[m$t2 >= m$t1 & m$t1>0,]                                              # add conditions: end of follow up (t2) needs                                                              to be after start of trial (t1), t1 must be larger than 0. 

# Add extra columns to matrix for C.OLE, E.OLE and Utility (U)
m$uc <- NA # 'Utility' C
m$ue <- NA # 'Utility' E
m$U <- NA # Overall utility weighted by the power 

# For loop
for(i in 1:nrow(m)) {
  m$uc[i] <- OLE(N = m$N[i], hE = m$hE[i], hC = m$hC[i], t1 = m$t1[i], t2 = m$t2[i], t_star = m$t_star[i])$'Overall life expectancy E not superior'
m$ue[i] <- OLE(N = m$N[i], hE = m$hE[i], hC = m$hC[i], t1 = m$t1[i], t2 = m$t2[i], t_star = m$t_star[i])$'Overall life expectancy E superior'
  m$U[i] <- OLE(N = m$N[i], hE = m$hE[i], hC = m$hC[i], t1 = m$t1[i], t2 = m$t2[i], t_star = m$t_star[i])$'Utility'
}


# Outcome 
m[m$U == max(m$U),] 
```


# Scenario 4
Experimental treatment is better then the control treatment (thus, lower hazard).
A small difference. N is relatively small (300). 

```{r}
# Data input
hE <-0.2
hC <- 0.25
t_star <-10
N <- 300

# Create matrix with possible combinations of t1 and t2
m <- expand.grid(hE=hE, hC=hC, t_star=t_star, N=N, t1=t1.seq, t2=t2.seq)    # t2 = column name, t2seq = content
m <- m[m$t2 >= m$t1 & m$t1>0,]                                              # add conditions: end of follow up (t2) needs                                                              to be after start of trial (t1), t1 must be larger than 0. 

# Add extra columns to matrix for C.OLE, E.OLE and Utility (U)
m$uc <- NA # 'Utility' C
m$ue <- NA # 'Utility' E
m$U <- NA # Overall utility weighted by the power 

# For loop
for(i in 1:nrow(m)) {
  m$uc[i] <- OLE(N = m$N[i], hE = m$hE[i], hC = m$hC[i], t1 = m$t1[i], t2 = m$t2[i], t_star = m$t_star[i])$'Overall life expectancy E not superior'
m$ue[i] <- OLE(N = m$N[i], hE = m$hE[i], hC = m$hC[i], t1 = m$t1[i], t2 = m$t2[i], t_star = m$t_star[i])$'Overall life expectancy E superior'
  m$U[i] <- OLE(N = m$N[i], hE = m$hE[i], hC = m$hC[i], t1 = m$t1[i], t2 = m$t2[i], t_star = m$t_star[i])$'Utility'
}

# Outcome 
m[m$U == max(m$U),] 
```


# Scenario 5
Experimental treatment is not better. 
A relatively large difference. Large N. 
```{r}
# Data input
hE <-0.3
hC <- 0.1
t_star <-10
N <- 3000

# Create matrix with possible combinations of t1 and t2
m <- expand.grid(hE=hE, hC=hC, t_star=t_star, N=N, t1=t1.seq, t2=t2.seq)    # t2 = column name, t2seq = content
m <- m[m$t2 >= m$t1 & m$t1>0,]                                              # add conditions: end of follow up (t2) needs                                                              to be after start of trial (t1), t1 must be larger than 0. 

# Add extra columns to matrix for C.OLE, E.OLE and Utility (U)
m$uc <- NA # 'Utility' C
m$ue <- NA # 'Utility' E
m$U <- NA # Overall utility weighted by the power 

# For loop
for(i in 1:nrow(m)) {
  m$uc[i] <- OLE(N = m$N[i], hE = m$hE[i], hC = m$hC[i], t1 = m$t1[i], t2 = m$t2[i], t_star = m$t_star[i])$'Overall life expectancy E not superior'
m$ue[i] <- OLE(N = m$N[i], hE = m$hE[i], hC = m$hC[i], t1 = m$t1[i], t2 = m$t2[i], t_star = m$t_star[i])$'Overall life expectancy E superior'
  m$U[i] <- OLE(N = m$N[i], hE = m$hE[i], hC = m$hC[i], t1 = m$t1[i], t2 = m$t2[i], t_star = m$t_star[i])$'Utility'
}

# Outcome 
m[m$U == max(m$U),] 
```

```{r}
    # grey = E is not superior
    # green = E is superior
    # red = utility 

# All 3 combined
plot3d(x = m$t1, y = m$t2, z = m$uc, col = 'grey') + plot3d(x = m$t1, y = m$t2, z = m$ue, add = T, col = 'green') + plot3d(x = m$t1, y = m$t2, z = m$U, add = T, col = 'red')
 
# Only utility
plot3d(x = m$t1, y = m$t2, z = m$U, col = 'red')
```


# Scenario 6
Experimental treatment is not better. 
But only a small difference. Large N. 
```{r}
# Data input
hE <-0.3
hC <- 0.25
t_star <-10
N <- 300

# Create matrix with possible combinations of t1 and t2
m <- expand.grid(hE=hE, hC=hC, t_star=t_star, N=N, t1=t1.seq, t2=t2.seq)    # t2 = column name, t2seq = content
m <- m[m$t2 >= m$t1 & m$t1>0,]                                              # add conditions: end of follow up (t2) needs                                                              to be after start of trial (t1), t1 must be larger than 0. 

# Add extra columns to matrix for C.OLE, E.OLE and Utility (U)
m$uc <- NA # 'Utility' C
m$ue <- NA # 'Utility' E
m$U <- NA # Overall utility weighted by the power 

# For loop
for(i in 1:nrow(m)) {
  m$uc[i] <- OLE(N = m$N[i], hE = m$hE[i], hC = m$hC[i], t1 = m$t1[i], t2 = m$t2[i], t_star = m$t_star[i])$'Overall life expectancy E not superior'
m$ue[i] <- OLE(N = m$N[i], hE = m$hE[i], hC = m$hC[i], t1 = m$t1[i], t2 = m$t2[i], t_star = m$t_star[i])$'Overall life expectancy E superior'
  m$U[i] <- OLE(N = m$N[i], hE = m$hE[i], hC = m$hC[i], t1 = m$t1[i], t2 = m$t2[i], t_star = m$t_star[i])$'Utility'
}

# Outcome 
m[m$U == max(m$U),] 
```

Plots
```{r}
    # grey = E is not superior
    # green = E is superior
    # red = utility 

# All 3 combined
plot3d(x = m$t1, y = m$t2, z = m$uc, col = 'grey') + plot3d(x = m$t1, y = m$t2, z = m$ue, add = T, col = 'green') + plot3d(x = m$t1, y = m$t2, z = m$U, add = T, col = 'red')
 
# Only utility
plot3d(x = m$t1, y = m$t2, z = m$U, col = 'red')
```


# Scenario 7 
Experimental treatment is not better. 
A relatively large difference. Small N. 
```{r}
# Data input
hE <-0.3
hC <- 0.1
t_star <-10
N <- 300

# Create matrix with possible combinations of t1 and t2
m <- expand.grid(hE=hE, hC=hC, t_star=t_star, N=N, t1=t1.seq, t2=t2.seq)    # t2 = column name, t2seq = content
m <- m[m$t2 >= m$t1 & m$t1>0,]                                              # add conditions: end of follow up (t2) needs                                                              to be after start of trial (t1), t1 must be larger than 0. 

# Add extra columns to matrix for C.OLE, E.OLE and Utility (U)
m$uc <- NA # 'Utility' C
m$ue <- NA # 'Utility' E
m$U <- NA # Overall utility weighted by the power 

# For loop
for(i in 1:nrow(m)) {
  m$uc[i] <- OLE(N = m$N[i], hE = m$hE[i], hC = m$hC[i], t1 = m$t1[i], t2 = m$t2[i], t_star = m$t_star[i])$'Overall life expectancy E not superior'
m$ue[i] <- OLE(N = m$N[i], hE = m$hE[i], hC = m$hC[i], t1 = m$t1[i], t2 = m$t2[i], t_star = m$t_star[i])$'Overall life expectancy E superior'
  m$U[i] <- OLE(N = m$N[i], hE = m$hE[i], hC = m$hC[i], t1 = m$t1[i], t2 = m$t2[i], t_star = m$t_star[i])$'Utility'
}

# Outcome 
m[m$U == max(m$U),] 
```

Plots
```{r}
    # grey = E is not superior
    # green = E is superior
    # red = utility 

# All 3 combined
plot3d(x = m$t1, y = m$t2, z = m$uc, col = 'grey') + plot3d(x = m$t1, y = m$t2, z = m$ue, add = T, col = 'green') + plot3d(x = m$t1, y = m$t2, z = m$U, add = T, col = 'red')
 
# Only utility
plot3d(x = m$t1, y = m$t2, z = m$U, col = 'red')
```


# Scenario 8 
Experimental treatment is not better. 
But only a small difference. Small N. 
```{r}
# Data input
hE <-0.25
hC <- 0.2
t_star <-10
N <- 300

# Create matrix with possible combinations of t1 and t2
m <- expand.grid(hE=hE, hC=hC, t_star=t_star, N=N, t1=t1.seq, t2=t2.seq)    # t2 = column name, t2seq = content
m <- m[m$t2 >= m$t1 & m$t1>0,]                                              # add conditions: end of follow up (t2) needs                                                              to be after start of trial (t1), t1 must be larger than 0. 

# Add extra columns to matrix for C.OLE, E.OLE and Utility (U)
m$uc <- NA # 'Utility' C
m$ue <- NA # 'Utility' E
m$U <- NA # Overall utility weighted by the power 

# For loop
for(i in 1:nrow(m)) {
  m$uc[i] <- OLE(N = m$N[i], hE = m$hE[i], hC = m$hC[i], t1 = m$t1[i], t2 = m$t2[i], t_star = m$t_star[i])$'Overall life expectancy E not superior'
m$ue[i] <- OLE(N = m$N[i], hE = m$hE[i], hC = m$hC[i], t1 = m$t1[i], t2 = m$t2[i], t_star = m$t_star[i])$'Overall life expectancy E superior'
  m$U[i] <- OLE(N = m$N[i], hE = m$hE[i], hC = m$hC[i], t1 = m$t1[i], t2 = m$t2[i], t_star = m$t_star[i])$'Utility'
}

# Outcome 
m[m$U == max(m$U),] 
```

Plots
```{r}
    # grey = E is not superior
    # green = E is superior
    # red = utility 

# All 3 combined
plot3d(x = m$t1, y = m$t2, z = m$uc, col = 'grey') + plot3d(x = m$t1, y = m$t2, z = m$ue, add = T, col = 'green') + plot3d(x = m$t1, y = m$t2, z = m$U, add = T, col = 'red')
 
# Only utility
plot3d(x = m$t1, y = m$t2, z = m$U, col = 'red')
```



